---
title: "Final Project Exploratory Data Analysis"
author: "Aidan Frederick, Ben Walter, Kyle Maher"
date: today
format:
  pdf:
    self-contained: true
documentclass: article
execute:
  warning: false
  message: false
header-includes:
  - \usepackage{pdfpages}
---

<!-- Directions: 
Conduct exploratory data analysis; visualize trends, seasonality, and stationarity tests. Visual and statistical exploration with a written summary. -->

# Packages
```{r setup}
library(fpp3) # Time Series Plots and Forecasting
library(corrplot) # Correlation Plot
library(forecast) # BoxCox Transformation
library(scales) # Log Scale Tick Lables
set.seed(280)  # Reproducibility
```

# Setup 
### Read Data
```{r read_data}
data <- readRDS("../data/loaded/delhi_filter.rds")
df <- data

df_monthly <- data %>%
  select(PM2.5) %>%
  index_by(month = yearmonth(datetime)) %>%
  summarise(PM2.5 = median(PM2.5, na.rm = TRUE))

df_monthly %>%
  autoplot(PM2.5)

df_weekly <- data %>%
  select(PM2.5) %>%
  index_by(week = yearweek(datetime)) %>%
  summarise(PM2.5 = median(PM2.5, na.rm = TRUE))

df_weekly %>%
  autoplot(PM2.5)

df_daily <- data %>%
  select(PM2.5) %>%
  index_by(date = as.Date(datetime)) %>%
  summarise(PM2.5 = median(PM2.5, na.rm = TRUE))

df_daily %>%
  autoplot(PM2.5)

```



# Recomended Limit for PM2.5
```{r PM2.5_limts}
df %>%
  index_by(date = as.Date(datetime)) %>%
  summarise(PM2.5 = mean(PM2.5, na.rm = TRUE)) %>%
  autoplot(PM2.5) +
  geom_hline(yintercept = 35, color = "red", linetype = "dashed") +
  geom_hline(yintercept = 12, color = "blue", linetype = "dashed")

```

The World Health Organization recomends PM2.5 levels below $35ug/m^3$ for a daily average and below $12ug/m^3$ for a yearly average. The daily and yearly recomended levels are plotted in red and blue respectively. Only during the summer months are the PM2.5 levels below the WHO recomended daily limit.


# Seasonality

```{r seasonality}

gg_season(df, PM2.5, period = "1 year") +
  labs(title = "Yearly Seasonality")

gg_season(df, PM2.5, period = "1 month") +
  labs(title = "Monthly Seasonality")

gg_season(df, PM2.5, period = "1 week") +
  labs(title = "Weekly Seasonality")

gg_season(df, PM2.5, period = "1 day") +
  labs(title = "Daily Seasonality")

```



# Parameter Correlations

```{r correlation_matrix}

M <- cor(x = df[ , c(2:10)],
         y = df[ , c(2:10)],
         use = "na.or.complete")
corrplot(M)

```
High concentrations of CO, NO, NO2, NOx, PM2.5, and PM10 generally occur at the same time. High concentrations of Benzene, Toluene, and Xylene generally occur at the same time.

# PM2.5 as a Stationary Time Series
```{r stationarity}

df$logPM25 <- log(df$PM2.5) %>%
    difference(24)

# Box-Cox Transformation
BoxCox.lambda(df$PM2.5)
df$BCPM25 <- BoxCox(df$PM2.5, BoxCox.lambda(df$PM2.5)) %>%
    difference(24)

df %>% autoplot(logPM25)
df %>% autoplot(BCPM25)
```


Differencing PM2.5 centers the data, and using a log transform centers the data, giving a stationary time series. However, using a Box-Cox transform creates a more stationary time series than the log transform. This method of transformation is likely the better option if we were to move forward with a model choice that requires stationarity.

# Check for Constant Variance

```{r variance_check}

delhi_stat <- data.frame(logvar = numeric(10),
                         boxvar = numeric(10))
for (i in 1:10){
    delhi_stat[i,1] <- var(df$logPM25[
        floor(nrow(df)*0.1*(i-1)):floor(nrow(df)*0.1*i)
        ], na.rm = TRUE)

    delhi_stat[i,2] <- var(df$BCPM25[
        floor(nrow(df)*0.1*(i-1)):floor(nrow(df)*0.1*i)
    ], na.rm = TRUE)
}

delhi_stat

plot(seq(1,10),delhi_stat$logvar, ylim = c(0,1))
```
The variance changes without transformation.

```{r boxcox_variance_plot}
plot(seq(1,10),delhi_stat$boxvar, ylim = c(0,9))
```
The Box-Cox transformation results in stable variances.


# World Health Organization Recomended Limits:
CO: ≤9 ppm  
PM₂.₅: ≤15 µg/m³  
PM₁₀: ≤45 µg/m³  
NO₂: ≤25 µg/m³  
Benzene: As low as possible (<5 µg/m³)  
Toluene: ≤260 µg/m³  
Xylene: ≤870 µg/m³  


# Scaled Parameters by Recomended Limits
```{r scaled_boxplots}

df_scaled <- df %>%
  select(-NOx, -NO) %>%
  mutate(
    PM2.5 = PM2.5 / 15,
    PM10 = PM10 / 45,
    CO = CO / 9,
    NO2 = NO2 / 25,
    Benzene = Benzene / 1,
    Toluene = Toluene / 260,
    Xylene = Xylene / 870
  ) %>%
  pivot_longer(c(2:8))

ggplot(data = df_scaled) +
    aes(x = name, y = value, color = name) +
    geom_boxplot() +
    scale_y_log10(labels = label_number()) +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.6)) +
    labs(
      x = NULL,
      y = "Multiple of Recomended Limit",
      title = "Air Quality Parameters by Recomended Limit"
    )
```

# Conclusion
TBD
